---
title: "R Tutorial for PMED Undergraduate Workshop"
author: "Xinyi Li, SAMSI"
date: "10/22/2018"
output:
  pdf_document: default
  html_notebook: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## About R

R is a free software environment for statistical computing and graphics:

* a different implementation of S developed at Bell Lab;  
* provides a wide variety of statistical and graphical techniques, and is highly extensible;  
* open source;  
* powerful IDE (integrated development environment), such as Rstudio.  

## Install R

1. Download the most recent version of [R](https://www.r-project.org/). The [R FAQs](https://cran.r-project.org/faqs.html) and the [R Installation and Adminstration Manual](https://cran.r-project.org/doc/manuals/R-admin.html) contain detailed instructions for installing R on various platforms (Linux, OS X, and Windows being the main ones).

2. Start the R program; on Windows and OS X, this will usually mean double-clicking on the R application, on UNIX-like systems, type "R" at a shell prompt. 

3. As a first step with R, start the R help browser by tying `help.start()` in the R command window. For help on any function, e.g. the "mean" function, type `?mean`.

## Install RStudio

1. Go to [RStudio](www.rstudio.com) and click on the "Download RStudio" button.

2. Click on "Download RStudio Desktop."

3. Click on the version recommended for your system, or the latest Windows version, and save the executable file. Run the .exe file and follow the installation instructions.

## Data types

We can use variables without definition in advance.

### Numbers

```{r}
num = 3.14
print(num)
print(num + 1)
print(typeof(num))
num.int = as.integer(num);
print(num.int)
print(typeof(num.int))
```

We can use `R` as a calculater, e.g. $2*2$, $\log(2)$, $\sqrt{2}$, $2^3$.

```{r}
x = 2
print(x * 2)
print(log(x))
print(sqrt(x))
print(x ^ 3)
print(x ** 3)
```

### Array

```{r}
y = 10:12
print(y)
z = c(1, 3, 5)
print(z)
print(z[1])
matrix = cbind(y, z)
print(matrix)
print(class(matrix))
print(matrix[2,])
print(matrix[c(2, 3),])
print(matrix[, 2])
```

### Data frame

```{r}
df = data.frame(y = y, z = z)
print(df)
print(class(df))
print(df$y)
print(df$z)
```


## Packages

### How to install a package

#### Install from source

Download the add-on `R` package, for example, "tidyverse", put it in the directory "/data/Rpackages", and install the package using the command:

```{r, eval = F, echo = T}
install.packages("tidyverse", lib = "/data/Rpackages")
```

#### Install from repository
Vast array of packages are available at the [Comprehensive R Archive Network (CRAN)](https://cran.r-project.org/) and [BioConductor](https://www.bioconductor.org/) repositories. Both CRAN and BioConductor are open source, well structured, tested and operating. While both repositories provide abundant packages covering various data analysis tasks, BioConductor is more focused on providing tools for the analysis of high-throughtput genomic data. In addition, there are slight differences in the command for package installation. 

* Install from CRAN (e.g. R package "tidyverse"):
```{r, eval = F, echo = T}
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
```

* Install from BioConductor (e.g. R package "dada2"):
```{r}
## try http:// if https:// URLs are not supported
source("https://bioconductor.org/biocLite.R")
biocLite()
```
```{r, eval = F, echo = T}
biocLite("dada2")
```

### How to load functions from a package
Type the following command in R console to load the package.
```{r, eval = T, echo = TRUE}
library(tidyverse)
library(dada2)
```


## Basic plots

### Use of "plot" function

```{r, eval = T, echo = TRUE}
x = 1:1000/10
y = sin(x)
print(head(x))
print(tail(y))
plot(x, y)
plot(x, y, type = "l")
plot(x[1:20], y[1:20], type = 'b', col = 'red', lwd = 5, ylim = c(0, 2))
```

### Use of "hist" function

```{r, eval = T, echo = TRUE}
set.seed(2018)
data = rnorm(1000)
plot(data, type = 'l')
hist(data)
```

Exercise: Use `hist()` and change parameters to generate the figure as follow.

```{r, eval = T, echo = FALSE}
hist(data, breaks = 50, freq = FALSE)
```

## Importing Data

Download data from https://github.com/johnnardini/SAMSI_PMED_UG, both "patient1.csv" and "patient_exp_data.zip". Upzip the zip file.

* Importing a single file

```{r, eval = T, echo = TRUE}
pati1 = read.csv("patient1.csv", header = TRUE)
print(class(pati1))
print(dim(pati1))
print(head(pati1))
```

* Importing multiple files

```{r, eval = T, echo = TRUE}
Data = NULL
for (i in 1:100) {
  fileName = paste0("patient_exp_data/patient", i, ".csv")
  dat = read.csv(fileName)
  dat$patient = i
  Data = rbind(Data, dat)
}
print(dim(Data))
print(head(Data))
print(tail(Data))
```
